---
title: "Gross Domestic Product Analysis"
date: "10-03-2017"
output:
  html_document:
    keep_md: yes
  pdf_document: default
authors: Ivelin Angelov, Laura Bishop, Ethan Graham, Scott Gozdzialski
---

# Introduction
The global economy is accelerating at a rapid pace.
The World Bank, which is the world's largest development institution, looks at influencing factors like climate change, conflict, food security, education, agriculture, finance, and trade.
Questions are being asked about Gross Domestic Product (GDP) and Income categories for 189 countries in which there is sufficient data to analyze. 

The data source for this analysis is from the World Bank's Education Statistics (http://datatopics.worldbank.org/education/) web site for basic information on a country, including income grouping, date of last census data, systems of trade, government accounting approach, and many others basic data points except GDP. The GDP information, also from the World Bank, is not a part of the Education Statistics information collection and dissemination. 

It is important to note that the World Bank's Education Statistics (EDSstat) does differentiate between countries that participate in the Organization of Economic Cooperation and Development (OECD). The OECD was founded in 1960 and contains 35 member countries today, including the United States, United Kingdom, and Germany.
The goal of this organization is to help developing countries create and sustain economic prosperity.
In this analysis, the  "Income Group" used in the ranking of GDP for a country does contain an indicator of whether a high income country is part of the OECD.<br/>

This analysis takes a closer look at the relationship between GDP and Income Groups.

## 1. With the merged data frame how many IDs matched?

After merging the Data from the EDStats with the GDP data, it is expected that some of the row will not perfectly match up.  Looking at the data from both sets and merging them together we end up with some data points that do not have information in the GDP, so we did not include those values in the data. So how many use full observations do we have?

```{r}
# Load the merged tidy data file
Data = read.table(file.path("data", "merged_data.csv"), header = TRUE, encoding='UTF-8')
cat('There are', nrow(Data), 'rows of clean merged data for further analysis.')
```


## 2. With the the data frame in decending order by GDP(United States last) what is the 13th value?
So, you may be wondering where different countries lie within this list of countries or what country lies in which position.
Here is an example of which county lie at the thirteenth position if we order the data in descending order.
This lists row in unordered data frame, three letter country code, country name, GDP in millions of US dollars, and finally which Organization of Economic Cooperation and Development group it belongs.

```{r}
# reorder by GDP
Data = Data[order(Data$GDP),]
# print 13-th
cat('The 13th country in the sorted dataset is:', as.character(Data$CountryName[13]))
```

## 3. What are the average GDP rankings for the "High Income: OECD" and "High Income:nonOECD" groups?

We have talked about different groups as classified by the Organization of Economic Cooperation and Development (OECD).
There are five different groups.
They are the high income and member of the OECD, the high income and non-member OECD, the upper middle income, the middle income, and the low income groups.
Now for us to look if there is a difference between being a member of the OECD makes a difference the only way we can do that is by comparing the average incomes of the high income OECD to the high income nonOECD, since they are the only groups with a counterpart.<br/> 

```{r}
Data$GDP.rankings = rank(Data$GDP)
OECD_rankings = Data$GDP.rankings[Data$Income.Group == "High income: OECD"]
cat('The "High income: OECD" group has an average GDP of', mean(OECD_rankings),'millions in US dollars.')

nonOECD_rankings = Data$GDP.rankings[Data$Income.Group == "High income: nonOECD"]
cat('The high income nonOECD group had an average GDP of', mean(nonOECD_rankings), 'millions in US dollars.')
```
You can see the OECD group has a higher average GDP.

## 4. Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.

It is hard to visualize all the data of the countries we have in our data frame in our minds.
To help with this we will plot that different groups in a chart below.
```{r message=F}
# Load needed packages
require(ggplot2)
# Allows for setting up the ggplot
require(scales) 
# For conclusion
require (Hmisc) 
# For histograms in conclusion
require (lattice) 

# Gets rid of the exponential numbers on the Y axis
options(scipen=10000)  

# Sets up GGPLOT2 scatter plot
ggplot (Data, aes(x=Data$Income.Group, y=Data$GDP)) + 
  # Sets up Y axis with labels and range
  scale_y_continuous(
    name="GDP -- In USD$ Millions",
    labels= scales::comma,
    expand = c(0,0),
    limits=c(0, 20000000),
    breaks=seq(0,18000000, 2000000)
    ) + 
  # Sets up x axis for abbreviates of categorical variable to prevent overlap and promote readability
  scale_x_discrete(
    "Income Group",
    labels = c("Low income" = "LI", "High income: nonOECD" = "nHI", "High income: OECD" = "oHI", "Lower middle income" = "LMI", "Upper middle income" = "UMI")
  ) + 
  # Sets up aspect ratio so consistent with each output
  theme(aspect.ratio = 2/1) + 
  # Sets up colored points
  geom_point(aes(color = Data$Income.Group)) +  
  # Changes legend title
  scale_color_discrete (name="Income Group")
```

## 5. Cut the GDP rankings into 5 seperate quantile groups. Making a table versus income group. How many countries are "lower middle income" but within the 38 nations with the highest GDP?
Looking at the chart above we can see that the groups are not built on GDP alone.
The selection into the high income and member of the OECD, the high income and non-member OECD, the upper middle income, the middle income, and the low income groups come from many variables.
With this knowledge, we can see that some members of the lower income groups may have the higher GDP then members of the high income group.
So, how many of the countries in the top 38 GDPs are actual in the lower middle income?  

```{r message=F}
# Split the rankings to 5 quantile groups
OrderData <- Data[order(Data$GDP.rankings), c(1:4)]
colNames <- c("Country Name", "Country Code", "GDP in Millions USD$", "Income Group")
# Load the thml table package
require('htmlTable')
# Print a table
htmlTable(
  OrderData[c('CountryName', 'CountryCode', 'GDP', 'Income.Group')],
  caption = "Ranking in Income Groups"
)
top38 <- tail(Data, n = 38)
# An answer to the question: How many countries are "lower middle income" but within the 38 nations with the highest GDP
cat('There are', NROW(top38[top38$Income.Group == 'Lower middle income', ]), 'countries "lower middle income" but within the 38 nations with the highest GDP')

```
## Conclusion
There are many factors that influence a country's GDP and overall prosperity.
This data analysis effort looks at GDP and the categories of Income Group as provided by the World Bank data.
Many things were looked at in this analysis of Gross Domestic Products of different countries around the world.
This took multiple steps to get here.

The first thing we had to do was get the data in a clean a usable format.
The cleaning and merging of the different dataframes left us with some incomplete records so we had to determine the number of usable records before we could go further into this study, we had 189 usable records.<br/>

It was then time to consider specific question, we wanted to look at a specific data point from the data when ordered in a descending order, finding the 13th nation from the top when sorted in descending order is St. Kitts and Nevis.
Afterwards, we wanted to see amongst the OECD and nonOECD groups was there a difference in their GDP, where we say the OECD high income group has a higher average GDP than the nonOECD high income group.
We then wanted to visualized the data.
Finally, after seeing the data we were curious about how strong GDP is in the OECD scoring and checked the number of nations in the top 38 GDPs that were ranked in the OECD lower middle income group, which is 5.
