---
title: "Gross Domestic Product Analysis"
output:
  html_document:
    keep_md: true
---
```{r echo=FALSE}
Data = read.table(file.path("data", "merged_data.csv"), header = TRUE, encoding='UTF-8')
```

#Introduction
The global economy is accelerating at a rapid pace.  The World Bank, which is the world's largest development institution, looks at influencing factors like climate change, conflict, food security, education, agriculture, finance, and trade.  Questions are being asked about Gross Domestic Product (GDP) and Income categories for 190 countries in which there is sufficient data to analyze. 

The data source for this analysis is from the World Bank's Education Statistics (http://datatopics.worldbank.org/education/) web site for basic information on a country, including income grouping, date of last census data, systems of trade, government accounting approach, and many others basic data points except GDP. The GDP information, also from the World Bank, is not a part of the Education Statistics information collection and dissemination. 

It is important to note that the World Bank's Education Statistics (EDSstat) does differentiate between countries that participate in the Organization of Economic Cooperation and Development (OECD). The OECD was founded in 1960 and contains 35 member countries today, including the United States, United Kingdom, and Germany.  The goal of this organization is to help developing countries create and sustain economic prosperity.  In this analysis, the  "Income Group" used in the ranking of GDP for a country does contain an indicator of whether a high income country is part of the OECD. 

This analysis takes a closer look at the relationship between GDP and Income Groups.



###1. With the merged data frame how many IDs matched?

When we merged the Educational data with the GDP data we were bound to have some that did not match.
The number that we ended up mathing together is listed below.
```{r}
nrow(Data)
```
###2. With the the data frame in decending order by GDP(United States last) what is the 13th value?

```{r}
# reorder by GDP
Data = Data[order(Data$GDP),]
# print 13-th
print(Data[c(13), ])
```

###3. What are the average GDP rankings for the "High Income: OECD" and "High Income:nonOECD" groups?

Comments
```{r}
Data$GDP.rankings = rank(Data$GDP)
OECD_rankings = Data$GDP.rankings[Data$Income.Group == "High income: OECD"]
print(mean(OECD_rankings))
nonOECD_rankings = Data$GDP.rankings[Data$Income.Group == "High income: nonOECD"]
print(mean(nonOECD_rankings)) 
```
###4. 1	Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.

Comments
```{r kable}
# install ggplot2 and scales in case its not
if("ggplot2" %in% rownames(installed.packages()) == FALSE){
  install.packages("ggplot2", repos = structure(c(CRAN = "http://cran.r-project.org")))
}
if("scales" %in% rownames(installed.packages()) == FALSE){
  install.packages("scales", repos = structure(c(CRAN = "http://cran.r-project.org")))
}
# load the packages
require(ggplot2)
require(scales)  # allows for setting up the ggplot

options(scipen=10000)  # gets rid of the exponential numbers on the Y axis

ggplot (Data, aes(x=Data$Income.Group, y=Data$GDP)) + # sets up GGPLOT2 scatter plot
  scale_y_continuous(name="GDP -- In USD$ Millions", labels= scales::comma, expand = c(0,0), limits=c(0, 20000000), breaks=seq(0,18000000, 2000000))  + #sets up Y axis with labels and range
  scale_x_discrete ("Income Group", labels = c("Low income" = "LI", "High income: nonOECD" = "nHI", "High income: OECD" = "oHI", "Lower middle income" = "LMI", "Upper middle income" = "UMI")) + # sets up x axis for abbreviates of categorical variable to prevent overlap and promote readability
  theme (aspect.ratio = 2/1) + #sets up aspect ratio so consistent with each output
  geom_point(aes(color = Data$Income.Group)) +  #sets up colored points
  scale_color_discrete (name="Income Group") #changes legend title
cat ("\n") # This pushes #5 below to the next line and allows the diagram to stand alone.
```   

###5. Cut the GDP rankings into 5 seperate quantile groups. Making a table versus income group.
###   How many countries are "lower middle income" but within the 38 nations with the highest GDP?

Comments
```{r}
# Split the rankings to 5 quantile groups
OrderData <- Data[order(Data$GDP.rankings), c(1:4)]
colNames <- c("Country Name", "Country Code", "GDP in USD$", "Income Group")
knitr::kable (OrderData[c('CountryName', 'CountryCode', 'GDP', 'Income.Group')], col.names = colNames, align = 'c', caption = "Ranking in Income Group")

top38 <- tail(Data, n = 38)
# this answer: How many countries are "lower middle income" but within the 38 nations with the highest GDP
print(NROW(top38[top38$Income.Group == 'Lower middle income', ]))

```

Conclusion