---
title: "Educational data vs Gross Domestic Product"
output:
  html_document:
    keep_md: true
---
```{r echo=FALSE}
Data = read.table(file.path("data", "merged_data.csv"), header = TRUE, encoding='UTF-8')
```

#Introduction
With the global market in flux the last few years and the rising cost of higher education, people are wondering if investment in higher education is worth the cost. Governments are also starting to wonder if the investment into higher education is justified when budget dollars could be spent on other needed projects. The DDS task force was created to provide guidance to governments and prospective students about the relationship between education and gross domestic product. The following analysis is performed based on Worldbank data on education and gross domestic product from 190 countries.

###1. With the merged data frame how many IDs matched?

When we merged the Educational data with the GDP data we were bound to have some that did not match.
The number that we ended up mathing together is listed below.
```{r}
nrow(Data)
```
###2. With the the data frame in decending order by GDP(United States last) what is the 13th value?

```{r}
# reorder by GDP
Data = Data[order(Data$GDP),]
# print 13-th
print(Data[c(13), ])
```

###3. What arethe average GDP rankings for the "High income: OECD" and "High Income:nonOECD" groups?

Comments
```{r}
Data$GDP.rankings = rank(Data$GDP)
OECD_rankings = Data$GDP.rankings[Data$Income.Group == "High income: OECD"]
print(mean(OECD_rankings))
nonOECD_rankings = Data$GDP.rankings[Data$Income.Group == "High income: nonOECD"]
print(mean(nonOECD_rankings))
```
###4. 1	Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.

COmments
```{r kable}
# install ggplot2 in case its not
if("ggplot2" %in% rownames(installed.packages()) == FALSE){
  install.packages("ggplot2", repos = structure(c(CRAN = "http://cran.r-project.org")))
}
# load the package
require(ggplot2)
qplot(Data$GDP, Data$GDP.rankings, color = Data$Income.Group)
```
###5. Cut the GDP rankings into 5 seperate quantile groups. Making a table versus income group.
###   How many countries are "lower middle income" but within the 38 nations with the highest GDP?

COmments
```{r}
# install dplyr in case its not
if("dplyr" %in% rownames(installed.packages()) == FALSE){
  install.packages("dplyr", repos = structure(c(CRAN = "http://cran.r-project.org")))
}
# load dplyr
library(dplyr)
# ntile is a function from dplyr
# ntile will split the rankings to 5 quantile groups
Data$QuantGroup <- ntile(Data$GDP.rankings, 5)
# I dont know what this means: Make a table versus Income.Group
#     is this is?
knitr::kable(Data[c('QuantGroup', 'Income.Group')])
top38 <- tail(Data, n = 38)
# this answer: How many countries are "lower middle income" but within the 38 nations with the highest GDP
print(NROW(top38[top38$Income.Group == 'Lower middle income', ]))
```

Conclusion