---
title: "Educational data vs Gross Domestic Product"
output: 
  html_document:
    keep_md: true
---
```{r}
if (.Platform$OS.type == 'unix') {
  data_path = 'data/'
} else {
  data_path = 'data\\'
}
Data = read.table(paste(data_path, "merged_data.csv", sep = ''), header = TRUE)
```

#Introduction
With the global market in flux of the last few years and the rising cost of higher education people are wondering if the investment is worth the cost.  At the same time governments are starting to wonder if the cost of investment into higher education is worth it when the limited budget dollars could be spent on other needed projects.  With this in mind we were asked to look into this problem and compare Education levels with Gross Domsetic Products to see if there was any relationship.  In particular we were asked the following questions.  To merge the data based on the country shortcode and How many IDs matched?  Sort the data frame in ascending order by GDP and what is the thirteenth country in the resulting dataframe?  What are the average GDP rankings for the "High income:OECD" and "High income: nonOECD" groups? To plot the GDP for all of the countries, and use ggplot2 to color you plots by income group. Finally to cut the GDP ranking into 5 seperate quantile groups, making a table versus income group. How many countries are lower middle income but are among the 38 nations with the highest GDP?

###1. With the merged data frame how many IDs matched?

When we merged the Educational data with the GDP data we were bound to have some that did not match.
The number that we ended up mathing together is listed below.
```{r}
nrow(Data)
```
###2. With the the data frame in decending order by GDP(United States last) what is the 13th value?

```{r}
# reorder by GDP
Data = Data[order(Data$GDP),]
# print 13-th
print(Data[c(13), ])
```

###3. What arethe average GDP rankings for the "High income: OECD" and "High Income:nonOECD" groups?

Comments
```{r,cache=TRUE}
nrow(EDSandGDPdata)
```
###4. Plot the GDP for all the countries, coloring the plot by income groups.

COmments
```{r,cache=TRUE}
nrow(EDSandGDPdata)
```
###5. Cut the GDP rankings into 5 seperate quantile groups. Making a table versus income group.
###   How many countries are "lower middle income" but within the 38 nations with the highest GDP?

COmments
```{r,cache=TRUE}
nrow(EDSandGDPdata)
```

COnclusion