---
title: "Makefile"
output: 
  html_document:
    keep_md: true
---
# This is the project's Makefile.
This file will download, clean, and merge the data files needed for the analysis of the "Case Study 1" prject.

#### <font color='red'>In order to Knit this file you need to have R Working Directory set to the root of the project.</font>

### Download all necessary for the project csv data files.
The files are downloaded from their original location as given by the assignment.
```{r download, message=F}
# Run the script in source/download.R
# All paths will be created according to the host OS requirements
source(file.path("source", "download.R"), verbose=FALSE)
```
### This cell will tidy the data files.
```{r clean_data, message=F}
source(file.path("source", "clean_gdp.R"))
source(file.path("source", "clean_income_groups.R"))
```
### Merge data the cleaned data files.
The cell bellow will merge the two tidy files: clean_gdp.csv and clean_income_groups.csv into merge.csv
```{r merge_data, message=F}
source(file.path("source", "merge.R"))
```
### Render the Analysis into a HTML file.
```{r render, message=F}
rmarkdown::render("Analysis.Rmd")
```